---
title: "Sampling The Imaginary"
author: "Tristan Mahr"
date: "July 18, 2016"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", collapse = TRUE)
```

## Preamble 

This is notebook of code I wrote while reading Chapter 3 of [_Statistical
Rethinking_](http://xcelab.net/rm/statistical-rethinking/).


## Standard medical testing example

You have a test to reports whether a patient is a vampire or not.

```{r}
# test is very accurate
p_test_given_vampire <- .95

# but false positives happen
p_test_given_mortal <- .01

# vampires are very rare
p_vampire <- .001

# Work out other helpful facts from these
p_mortal <- 1 - p_vampire
p_test <- p_mortal * p_test_given_mortal + p_vampire * p_test_given_vampire
```

Someone takes the test and it says yes. What are the chances that they are a
vampire?

We uses Bayes's Theorem. P(Vampire | Test) = P(Test | Vampire) * P(Vampire) / P(Test)

```{r}
p_vampire_given_test <- p_test_given_vampire * p_vampire / p_test
p_vampire_given_test
```

Only an `r round(p_vampire_given_test * 100, 1)`% chance of being a vampire,
given that they tested positive on the test.

But using Bayes's theorem is not the same as Bayesian inference. We could have
computed this using frequencies of events.

```{r}
# Suppose the test were administered 100,000 times.
n_tested <- 100000
n_mortal <- 99900
n_vampires <- 100
n_vampires_tested_yes <- 95
n_mortals_tested_yes <- 999
```

What are the chances of a yes being a vampire?

```{r}
n_vampires_tested_yes / (n_vampires_tested_yes + n_mortals_tested_yes)
95 / (999 + 95)
```

Bayesian inference is about probability over parameters. The parameters here are
well-defined.

*** 

## Globe tossing example

Globe tossed 9 times, and it lands on water 6 times. What is proportion of earth
is covered with water on the globe?

Grid approximation of posterior distribribution. Assuming all probabilities
are equally likely (`prior`). Try many probability values (`p_grid`) from 0 to
1. Measure the `likelihood` of the data at each probability. Divide by total
likelihood to get the posterior probability of each probability in the grid.

```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1 , 1000)
likelihood <- dbinom(6, size = 9, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
# Sampling from the posterior
samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)

plot(samples)
plot(density(samples, adjust = .5))
```

How to find the point with peak density

```{r}
# find the point at peak density
dens <- density(samples)
dens
plot(dens)
dens$x[which.max(dens$y)]
```


## Practice

### Hard

```{r}
library("tibble")
library("dplyr")
library("tidyr")
data(homeworkch3, package = "rethinking")

# 100 two-children families. 1: Child is a boy. 0: Girl.
df <- data_frame(
  Family = seq_along(birth1),
  Child1 = birth1,
  Child2 = birth2)
df

# Long format better reflects nesting of children in families
df_long <- df %>% 
  rename(`1` = Child1, `2` = Child2) %>% 
  gather(Child, Gender, -Family) %>% 
  readr::type_convert() %>% 
  arrange(Family, Child)
df_long

```

**3H1. Using grid approximation, compute the posterior distribution for the
probability of a birth being a boy. Assume a uniform prior probability. Which
parameter value maximizes the posterior probability?**

Basically, rework through the globe-tossing grid approximation's code with new
data.

```{r}
births <- df_long$Gender

p_grid <- seq(from = 0, to = 1, length.out = 1000)
prior <- rep(1 , 1000)

likelihood <- dbinom(sum(births), size = length(births), prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

p_grid[which.max(posterior)]
```

**3H2. Using the `sample` function, draw 10,000 random parameter values from the
posterior distribution you calculated above. Use these samples to estimate the
50%, 89%, and 97% highest posterior density intervals.**

```{r}
samples <- sample(p_grid, prob = posterior, size = 1e5, replace = TRUE)
plot(density(samples))

# HPDI returned in a data-frame with a column for probability level
tidy_hpdi <- function(xs, prob) {
  xs %>% 
    coda::as.mcmc() %>% 
    coda::HPDinterval(prob) %>% 
    as.data.frame %>% 
    rownames_to_column("Variable") %>% 
    as_data_frame() %>% 
    mutate(Level = prob) %>% 
    select(Level, Variable, lower, upper)
}

bind_rows(
  tidy_hpdi(samples, .5),
  tidy_hpdi(samples, .89),
  tidy_hpdi(samples, .97)
)
```

**3H3. Use `rbinom` to simulate 10,000 replicates of 200 births. You should end
up with 10,000 numbers, each one a count of boys out of 200 births. Compare the
distribution of predicted numbers of boys to the actual count in the data (111
boys out of 200 births).**

```{r}
replicates <- rbinom(n = length(samples), size = 200, prob = samples)
plot(density(replicates))
abline(v = sum(births))
```

**3H4. Now compare 10,000 counts of boys from 100 simulated first borns only
to the number of boys in the first births, `birth1`. How does the model look in
this light?**

The data don't fall squarely in the middle of the simulations. But they fall in
40% HPDI.

```{r}
replicates <- rbinom(n = length(samples), size = 100, prob = samples)
plot(density(replicates))
abline(v = sum(df$Child1))

sum(df$Child1)
tidy_hpdi(replicates, .4)
```

**3H5. The model assumes that sex of first and second births are independent.
To check this assumption, focus now on second births that followed female first
borns. Compare 10,000 simulated counts of boys to only those second births
that followed girls. To do this correctly, you need to count the number of
first borns who were girls and simulate that many births, 10,000 times. Compare
the counts of boys in your simulations to the actual observed count of boys
following girls. How does the model look in this light? Any guesses what is
going on in these data?**

The parents want the second child to be a boy. The number of boys born after
girls falls on the extreme edge of the simulations.

```{r}
n_girl_birth1 <- sum(birth1 == 0)
n_girl_birth1

boys_after_girls <- df %>% 
  filter(Child1 == 0, Child2 == 1) %>% 
  nrow
boys_after_girls

replicates <- rbinom(n = length(samples), size = n_girl_birth1, prob = samples)
plot(density(replicates))
abline(v = boys_after_girls)
```

It's possible that they want one of each gender. The number of boys born after
girls is also kind of skewed away from the simulations, but nowhere near as
extremely. Maybe because some parents are fine with two boys.

```{r}
n_boy_birth1 <- sum(birth1 == 1)

girls_after_boys <- df %>% 
  filter(Child1 == 1, Child2 == 0) %>% 
  nrow
girls_after_boys

boys_after_boys <- df %>% 
  filter(Child1 == 1, Child2 == 1) %>% 
  nrow
boys_after_boys

replicates <- rbinom(n = length(samples), size = n_boy_birth1, prob = samples)
plot(density(replicates))
abline(v = boys_after_boys)

```


```{r}
df %>% count(Child1, Child2)
```

