---
title: "06 - Information Criteria"
author: "Tristan Mahr"
date: "June 29, 2016"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", collapse = TRUE)
```


```{r packages}
# devtools::install_github("rmcelreath/rethinking")
library("dplyr", warn.conflicts = FALSE)
library("rstanarm")
library("ggplot2")
# options(mc.cores = parallel::detectCores())

# I don't load the rethinking package, but I use the data from it.
rethinking_info <- packageDescription(
  pkg = "rethinking", 
  fields = c("Version", "GithubUsername", "GithubRepo", 
             "GithubRef", "GithubSHA1"))
str(rethinking_info, give.attr = FALSE)
```


## 6.5 Using Information Criteria

Get the primate milk data. The example models will predict the kilocalories per
gram of milk using the mass of the mother (log kg) and the proportion of the
brain that is neocortex. 

```{r}
data(milk, package = "rethinking")
d <- milk %>% 
  as_data_frame %>% 
  filter(!is.na(neocortex.perc)) %>% 
  mutate(neocortex = neocortex.perc / 100,
         log_mass = log(mass)) %>% 
  select(clade, species, kcal.per.g, mass, log_mass, neocortex)
d %>% knitr::kable(digits = 3)
```

I fit the models using RStanArm, and not the functions provided by the
rethinking package, so I am not using the same priors as the book.

Fit the models.

```{r}
# Intercept only
m1 <- stan_glm(
  formula = kcal.per.g ~ 1,
  data = d,
  family = gaussian(),
  prior = normal(0, 1),
  prior_intercept = normal(.5, .5),
  prior_ops = prior_options(prior_scale_for_dispersion = .25)
)

# One predictor
m2 <- update(m1, . ~ neocortex)
m3 <- update(m1, . ~ log_mass)

# Two predictors
m4 <- update(m1, . ~ neocortex + log_mass)
```

Compare posterior to prior in each model.
 
```{r posterior-vs-prior}
# Want to have same colors for parameters across plots, so manually set colors
# so that intercept gets first color, sigma fourth color, etc.
d3_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", 
               "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf")
p1 <- posterior_vs_prior(m1) + scale_color_manual(values = d3_colors[c(1, 4)])
p2 <- posterior_vs_prior(m2) + scale_color_manual(values = d3_colors[c(1, 3:4)])
p3 <- posterior_vs_prior(m3) + scale_color_manual(values = d3_colors[c(1:2, 4)])
p4 <- posterior_vs_prior(m4) + scale_color_manual(values = d3_colors[c(1:4)])
cowplot::plot_grid(p1, p2, p3, p4)
```

Get the WAIC using the loo package.

I'm going to experiment with many-models, dataframe-of-models technique here.
That means doing some unconventional things with dataframes. 

Basically, I make a data-frame with one row per model and then do things to each
model, storing the results of those operations as new columns in the dataframe.
Specifically, I calculate the WAIC of the Stan models and refit the models
using the classical technique so I can get an AIC value to compare to the WAIC
value.

```{r}
library("tidyr")
library("purrr")

# Refit a stanreg model using lm
as_lm <- function(stan_model) {
  lm(stan_model$formula, stan_model$data )
}

# Get a dataframe summary from a loo::waic object
tidy_waic <- function(waic_fit) {
  to_keep <- c("waic", "se_waic", "elpd_waic", "se_elpd_waic", 
               "p_waic", "se_p_waic")
  ret <- waic_fit[to_keep]
  as_data_frame(ret)
}

# Create a dataframe with one-row per model
model_summary <- data_frame(StanFit = list(m1, m2, m3, m4)) %>% 
  mutate(
    # Use the formula from the model as a label
    Formula = map_chr(StanFit, ~ .x$formula %>% deparse),
    # Get the classical fit and deviance/AIC of classical fit
    ClassicalFit = map(StanFit, as_lm),
    Deviance = map_dbl(ClassicalFit, ~ logLik(.x) * -2),
    AIC = map_dbl(ClassicalFit, AIC),
    # Get the WAIC summary from the Stan fit of the model
    WAIC = map(StanFit, . %>% loo::waic() %>% tidy_waic)) %>% 
  # The WAIC is stored as a dataframe within each row. Unnest to promote the
  # columns from the nested dataframe
  unnest(WAIC)

# We need to exclude the list columns in order to print a formatted table 
model_summary %>% 
  select(-StanFit, -ClassicalFit) %>% 
  arrange(waic) %>% 
  knitr::kable(, digits = 2) 
```



```{r}

```


```{r}



```



